\Title

### 2013-01-23 synmut.tex
Text adapted to new figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-C5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
\noindent
Intrapatient HIV evolution is governed by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach high frequencies.  We analyze time
series of viral sequences from the \shankaregion~part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthusian) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
its involvement in RNA stem-loop structures.
Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
selective sweeps are less dense and hitch-hiking not as strong.

@@

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
skewed codon usage bias (CUB) patterns towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, while it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative. Their ability to hitchhike for extended times, which is a
core requirement for our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. This is a simple consequence of the fact
that (a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation, see
illustration in \FIG{fixp}.  Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction). They therefore represent
a simple way to investigate average properties of different classes of
mutations.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency (dashed lines), very few fix. This
observation is quantified in panels \FIG{fixp1} and \ref{fig:fixp2}, which
stratify the data of 7 (resp. 10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly two years and are lost much more frequently than expected. The long-time
probability of fixation versus extinction is shown as a function of the initial
frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). In contrast to synonymous
mutations, the nonsynonymous seem to follow more a less the neutral expectation
(blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8.pdf}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each (see methods). Repeating the above analysis separately on the
C2-V5 region studied above and the remainder of \env~ reveal striking
differences (see \FIG{fixp}). Within C2-V5, this data fully confirms the
observations made in the data set by \citet{shankarappa_consistent_1999} (red
line). In the remainder of \env, however, observed synonymous mutations behave
as if they were neutral (orange line).

%ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? (?? SOME!)
%DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: SUPPLEMENTARY?
%CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx
2~\text{\textperthousand}$). In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of them will be helix breakers). Furthermore, as a
complementary analysis, we split the synonymous mutations in the extended V1-V5
region further into conserved and variable regions and found that the biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness effect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, evidence for any contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, although codon-optimized HIV seems to perform better
{\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
at the macroevolutionary level. Second, within a single patient, we do not
observe any bias towards more human-like CUB in the synonymous mutations that
reach fixation rather than extinction. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.
\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
\label{fig:SHAPEA}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep{smith_hitch-hiking_1974}.

@@

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we perform computer simulations of
evolving viral populations under selection and rare recombination. For this
purpose, we use the recently published package FFPopSim, which includes a module
dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
analyze many combinations of parameters such as population size, recombination
rate, selection coefficient and density of escape mutations, and deleterious effect
of synonymous mutation.

The main result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and result in a dependence of the
fixation probability on initial frequency that is compatible with observations.
First of all, since neutral mutations are much more likely to rise to high
frequency than deleterious ones, the majority of the synonymous mutations needs
to be slightly deleterious observe a significant reduction of $P_\text{fix}$.
In order to further quantify the reduction in fixation probability, we look at
the difference between the neutral curve ($P_\text{fix}(\nu) = \nu$) and the
measured fixation probability and calculate its area (see inset of
\FIG{simfixpvar}). The minimal and maximal values for this area are zero
(neutral-like curve) and 0.5 (no fixation at all), respectively. The HIV data
correspond to an area under the diagonal of approximately 0.2 for synonymous
changes, and a very small area over the diagonal for nonsynonymous changes.
Various simulation curves are shown in \FIG{simfixpvar}. Then, in
\FIG{simsfig}, we explore the parameter space: the combinations that
yield areas close to the experimental result are shown.
The two crucial parameters that control the fixation probability are the
following: (a) the deleterious effects of hitchhikers compared to the beneficial
effects of escape mutants, and (b) the density of escape mutations. Intuitively,
a higher density of escape mutations (i.e., epitopes) enables a larger degree of
genetic draft, because escape mutations from different epitopes start to combine
and their effects add up. In \FIG{simheat1}, we show that this is indeed the
case in simulations.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
effect size of the synonymous alleles. This parameter also reduces synonymous
diversity (first inset). The depression found in HIV data is about 0.2.
(Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity.}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-C5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
\noindent
Intrapatient HIV evolution is governed by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach high frequencies in the viral population.  We analyze time
series of viral sequences from the \shankaregion~part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely reach fixation.
Simulations suggest that such synonymous mutations
have a (Malthusian) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles. We detect a negative correlation between the fixation of an allele and
its involvement in RNA stem-loop structures.
Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
selective sweeps are less dense and hitchhiking not as strong; this behaviour is
confirmed by extensive computer simulations.

@@

Escape mutations are strongly selected for their effect on the amino acid
sequence of the HIV proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative; their ability to hitchhike for extended times itself is
rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{fixp},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation.
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction). They therefore represent
a simple way to investigate average properties of different classes of
mutations.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency, very few fix
(panel~\ref{fig:aftsyn}); however, quite some nonsynonymous mutations reach
fixation (panel~\ref{fig:aftnonsyn}).
These observations are quantified in \FIG{fixp}, which stratifies the data
of 7 (resp. 10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly two years and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$
in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back
below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
\label{fig:aftsyn}}\\
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 8~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e. its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each (see methods). Repeating the above analysis separately on the
C2-V5 region studied above and the remainder of \env~ reveal striking
differences (see \FIG{fixp2}). Within C2-V5, this data fully confirms the
observations made in the data set by \citet{shankarappa_consistent_1999} (red
line). In the remainder of \env, however, observed synonymous mutations behave
neutrally (orange line).

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of them will be helix breakers). Furthermore, as a
complementary analysis, we split the synonymous mutations in the extended V1-V5
region further into conserved and variable regions and found that the biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
\label{fig:SHAPEA}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
effect size of the synonymous alleles (in HIV data is about $-0.2$).
This parameter also reduces synonymous
diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
normalized by the number of possible mutations (first inset).
(Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity.}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we perform computer simulations of
evolving viral populations under selection and rare recombination. For this
purpose, we use the recently published package FFPopSim, which includes a module
dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
pick many random combinations of the following four parameters: effect
size and rate of appearance of beneficial escape mutations, and effect size
and fraction of deleterious synonymous mutations.

The first result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and result in a dependence of the
fixation probability on initial frequency that is compatible with observations
(see \FIG{simfixpvar}).
The fixation probability of synonymous alleles decreases from the neutral
expectation to zero as their fitness effect worsens (\FIG{simfixpvar},
from $-0.0003$ in blue to $-0.003$ in red). Contextually,
the number of alleles that reach high frequencies, which is a proxy for
synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).

To assess the reduction in fixation probability quantitatively, we calculathe
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
changes, and essentially zero for nonsynonymous changes.

Among all simulations, we select the one that show a large depression in
fixation probability of synonymous mutations but, simulteneously, a high
synonymous diversity, as observed in HIV data. This selection imposes a strong
constraint on the parameters, in particular on the ones related to synonymous
changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \lesssim 0.002$, is required. This result fits well
the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
Moreover, it corresponds to {\it a priori} theoretical expectations:
(a) since high frequency bins are enriched in neutral mutations (it is easier for
them to hitchhike), the hallmark of deleterious mutations will be visible there
only if they are pervasive; (b) genetic draft stops working if the
deleterious effect size becomes comparable to the adaptive effect size of escape
mutations, because the double mutant has little or no fitness advantage.
